Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-11 paper/ch_potentials.tex
more wordsmithing
iterative cch pot and minor notation adjustments

% TODO Capital letters for arrays?

Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this shortest path problem.
In this section, we recall the classical algorithm and establish some notation that we will employ in the remainder of this paper.
Dijkstra's algorithm maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $d[v]$ is set to $\min(d[v], d[u] + w(uv))$.
If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
			$\Dist(P[x])$\;
		}
		$D[x]\leftarrow B[x]$\;
		\For{all up-edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}
% TODO iterative

@@

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
The forward search can skip nodes already settled by the backward search and vice versa.
However, this straightforward bidirectional A* still performs more work than a unidirectional A*.
Thus, on its own it is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

@@

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
The stopping criteria for each direction remain the same (unidirectional) as before.
However, often the search stops early because the queues are empty.

>>>>

Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this shortest path problem.
In this section, we recall the classical algorithm and establish some notation that we will employ in the remainder of this paper.
Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $D[s]$ is set to zero and the queue initialized with $s$.
Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
Once $t$ was settled, $D[t]$ contains the distance between $s$ and $t$.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\KwData{$S$: stack with nodes to compute distances}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwFunction{Push}{push}
\SetKwFunction{Pop}{pop}
\SetKw{Break}{break}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
  $y \gets x$\;
	\While{$D[y] = \bot$}{
    $S$.\Push{$y$}\;
		\If{$P[y] = \bot$}{
      \Break\;
		}
    $y \gets P[y]$
  }
  \While{$S$ not empty}{
    $y \gets S$.\Pop{}\;
		$D[y]\gets B[y]$\;
		\For{all up-edges $yz$ in $G^+_\ell$}{
                        $D[y]\gets\min\{D[y],w_\ell(yz)+D[z]\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}

@@

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
The forward search can skip nodes already settled by the backward search and vice versa.
However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
Thus, on its own it is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

@@

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
The stopping criteria for each direction remain the same (unidirectional) as before.
However, usually the search stops early because the queues are empty.
